Efficient and Scalable Bipartite Matching with Fast Beta Linkage (fabl)

### 2021-10-14 methods_paper/fabl-paper.tex
dont make biblio small.
references come before appendix.
addressed bekas comments

In this paper, we propose the \emph{fast beta linkage} method (\texttt{fabl}), which extends a recent Bayesian Fellegi Sunter model for increased efficiency and scalability. Specifically, we relax the one-to-one matching requirement of the Beta Record Linkage method of \citep{sadinle_bayesian_2017} and propose independent priors over the matching space. \textcolor{red}{We derive the associated conditional distributions and provide a Gibbs sampling algorithm for our proposed framework.}
\textcolor{red}{Our modified prior} allows us to employ hashing techniques that hasten calculations and reduce the overall computational complexity.  In addition, we are able to complete pairwise record comparisons over large datasets through parallel computing and reduce memory costs through a new technique called \emph{storage efficient indexing}. \textcolor{red}{We derive the associated conditional distributions and provide a Gibbs sampling algorithm for our proposed framework.}
\textcolor{red}{Moreover, we provide the computational complexity of \texttt{fabl}.} Through simulation studies and a case study of homicides from the El Salvadoran Civil War, we show that our method has markedly increased speed with minimal loss of accuracy. Finally, we discuss our work and future directions. 
\end{abstract}

@@

%	I have some proposed revisions for the first part of the paper. In the introduction, I reordered things and changed some topic sentences to highlight that the main contribution here is speed and scale. 
%	
%	Also, in the prior version, we didn't get to any of the proposed material until Section 4.3, which did not seem reasonable to me.
%	I know you recommended splitting the review of the literature between the Introduction and Section 2 on bipartite matching, but since we needed the mathemetical overview of FS and Sadinle, I decided to move all of the literature review into the introduction. I've seen this done in many papers. 
%	
%	Almost all of the content is still there, but it's moved around in a way that makes more sense to me. Let me know your thoughts! (But please avoid hard edits until we talk on Monday!)
%	
%	I have some additional edits in the later part of the paper about the partial Bayes estimate, computational complexity, and added detail for the case study. All of that is in blue as well.

	Record linkage is the task of identifying duplicate records across multiple data sources, often in the absence of a unique identifier \citep{christen_2012}. This is an increasingly important task in ``data cleaning,'' and is used for inferential and predictive analyses in fields such as statistics, computer science, machine learning, political science, economics, precision medicine, official statistics, and others. \textcolor{red}{In this paper, we consider bipartite record linkage, which merges two databases together that contain duplications across but not within the respective databases \citep{fellegi_theory_1969, jaro1989, Winkler1988, belin_1995, larsen_2001, liseo_2011,  herzog2007data, gutman_bayesian_2013, sadinle_bayesian_2017}.}
	%Much of the statistical literature focuses on bipartite record linkage \citep{fellegi_theory_1969, jaro1989, Winkler1988, belin_1995, larsen_2001, liseo_2011,  herzog2007data, gutman_bayesian_2013, sadinle_bayesian_2017}. 
	Many statistical record linkage methods are extensions of the seminal work of \cite{fellegi_theory_1969} and \cite{newcombe_automatic_1959}. Specifically, Fellegi and Sunter created comparison vectors for each pair of records in the data and independently classified those pairs as a match or a non-match using a likelihood ratio test. Recent work in the statistical literature has extended this approach for a wide variety of applications \citep{winkler1991application, fair2004generalized, wagner2014person, gill2003english}. Although the need to compute comparisons for all record pairs initially restricted this family of methods, recent work by \cite{enamorado2019using} used innovative hashing techniques to scale the approach to handle large datasets.

@@

	Alternatively, one can embed one-to-one matching requirements into the model specification itself, at an additional computational cost. \cite{larsen2005} employed a Metropolis-Hastings algorithm to only allow sampling matches that respected one-to-one assumptions, but such algorithms exhibit slow mixing due to the combinatorial nature of the constrained matching space. \cite{fortunato_2010} used simulated annealing to target the space of matches permitted under the one-to-one constraint, but the method was so computationally intensive it could only be used to link databases of less than 100 records. Most recently, \cite{sadinle_bayesian_2017} proposed the \emph{Beta Record Linkage} model (\texttt{BRL}), using an innovative prior over the space of bipartite matchings to strictly enforce one-to-one reqiurements throughout his Gibbs sampler. Additionally, he introduced a class of loss functions that allows for a flexible estimation of the linkage structure, such that the modeller can weight the relative importance of false positives and false negatives, and identify records pairings to be be decided through clerical review. Although it was shown to work on larger tasks than previous one-to-one methods, \texttt{BRL} becomes slow when working with larger datasets, and unusable when the number of pairwise record comparisons becomes too large.

	In this paper, we propose \emph{fast beta linkage} (\texttt{fabl}), which extends the the Beta Record Linkage model for increased efficiency and scalability. Specifically, we relax the one-to-one matching requirement of \texttt{BRL} and propose independent priors over the matching space, creating a ``many-to-one'' model for record linkage. This allows us to (1) employ hashing techniques that hasten calculations and reduce computational costs; (2) compute the pairwise record comparisons over large datasets via parallel computing; and (3) reduce memory costs through our proposed method of \emph{storage efficient indexing}. Throughout, we argue that even in cases where a bipartite matching is desired, our proposed approach (with a simple post-processing procedure) provides equivalently accurate estimation of the linkage structure and other parameters, more information through which to asses model misspecification, and greatly enhanced speed. Open source software to use \texttt{fabl} in \texttt{R} is available through Github at \textcolor{red}{BK: Add the URL even if it's private at the moment: \url{}}.

	\textcolor{red}{The remainder of the paper is as follows.} Section~\ref{sec:review-of_prior-work} reviews \textcolor{red}{the prior work of \cite{fellegi_theory_1969} and \cite{sadinle_bayesian_2017}.} In section~\ref{sec:fast-beta-linkage}, we propose our the \texttt{fabl} model specification. \textcolor{red}{Also, we derive the corresponding conditional distributions and provide a Gibbs sampler for posterior inference.}  Section \ref{bayes-estimate} \textcolor{red}{Brian: add in what you do in this section.}
	Section~\ref{sec:efficiency} \textcolor{red}{proposes our approach} for hashing and  and storage efficient indexing used to increase the speed of calculations and the scale of the linkage tasks we can undertake. In addition, we prove the computational complexity of \texttt{fabl}. Sections~\ref{sec:simulations} and \ref{sec:case-studies} demonstrate the speed and accuracy of \texttt{fabl} through simulation studies and a case study of homicides from the El Salvadoran Civil War. Section~\ref{discussion} concludes our paper with a discussion of directions for future work.

	% I think this is clear and has already been said. 
%	To recognize the lineage from the original \texttt{BRL} method, we name our method \emph{fast beta linkage} (\texttt{fabl}, pronounced ``fable'')

@@

In practice $P(\cdot \mid \Delta_{ij} = 1)$ and $P(\cdot \mid \Delta_{ij} = 0)$ are unknown to the modeler, and thus must be estimated. Under the Fellegi Sunter model, we assume both that comparison vectors are independent given their matching status, and that the match status of the record pairs are independent. This allows us to model 
the comparison data using mixture models in the following way:
\begin{align}
	&\Gamma_{ij} = \gamma_{ij} \mid \Delta_{ij} = 1 \stackrel{iid}{\sim} \mathcal{M}(\bm{m}), \\
	&\Gamma_{ij} = \gamma_{ij} \mid \Delta_{ij} = 0  \stackrel{iid}{\sim} \mathcal{U}(\bm{u}), \notag\\
	& \Delta_{ij}   \stackrel{iid}{\sim} \text{Bernoulli}(p), \notag
\end{align}
where $p$ is the proportion of records that match, and $\bm{m}$ and $\bm{u}$ are vectors of parameters for the matching and nonmatching record pairs. These unknown parameters are often estimated using the EM algorithm.

@@

Importantly, this independence means that we have relaxed the one-to-one requirement from \texttt{BRL}; our sampler does ensure that each record in \(\bm{X}_2\) can be matched to at most one record in \(\bm{X}_1\), but allows for the possibility that multiple records in \(\bm{X}_2\) match to the same record in \(\bm{X}_1\). This behavior may be desirable in its own right; \cite{Newcombe1962b} were interested in matching birth records to marriage certificates, a situation in which such ``many-to-one'' matchings are reasonable. In situations when a bipartite estimate is desired, we can use a simple post processing procedure shown in  section~\ref{bayes-estimate} to resolve illegal matchings.

@@

With the likelihood in this form, we can derive the full conditional
\[p(Z_j|\Gamma_{.j}, \Phi, \pi) \propto p(\Gamma_{.j}| Z_j, \Phi) P(Z_j|\pi)\]

\[\propto \left(\sum_{i=1}^{n_1}w_{ij}\bm{1}_{z_j = i} + \bm{1}_{z_j = n_1 + 1}\right)\left(\pi\sum_{i=1}^{n_1}\frac{1}{n_1}\bm{1}_{z_j = i} + (1-\pi)\bm{1}_{z_j = n_1 + 1}\right)\]
\[= \frac{\pi}{n_1}\sum_{i=1}^{n_1}w_{ij}\bm{1}_{z_j = i} + (1-\pi)\bm{1}_{z_j = n_1 + 1} \]
\[ \implies Z_j^{(s+1)} | \Phi, \Gamma, \pi \propto
\begin{cases} 
	\frac{\pi}{n_1}w_{ij}   & z_j \leq n_1; \\
	1-\pi &  z_j  = n_1 + 1 \\
\end{cases}\]

@@

To increase the speed of our Gibbs sampler and expand the scale of the linkage tasks we can undertake, we \textcolor{red}{propose} hashing methods similar to those used by \cite{enamorado2019using} in the creation of \texttt{fastlink}, a fast and scalable implementation of the Fellegi Sunter method. The key insight is to recognize that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. To make this more precise, let $h_1, \ldots, h_P$ denote the unique agreement patterns, and collect these unique patterns in the set $\mathcal{P} = \{h_1, \ldots, h_P\}$. Here, 
$P = |\mathcal{P}|$ is the total number of unique agreement
patterns; observe that \(P\) is bounded above by \(\prod_{f=1}^F L_f\), and that this bound does not depend on \(n_1\) or \(n_2\). In this context, the integers $\{1, \ldots, P\}$ serve as \emph{hashed values} that encode the same information as the orignial vectors themselves. Whenever possible, we conduct calculations over these \(P\) agreement
patterns, instead of the typical \(n_1 \times n_2\) record pairs.

@@

We have presented a method for Bayesian record linkage that is feasible for large datasets. In particular, our hashing procedure and model assumptions allow for a linkage procedure whose computational complexity does not scale with the size of the larger dataset, making this method particularly powerful in linking records when one datafile is
substantially smaller than the other.

In our case study, we included an exploration of how to conduct record linkage when modeling assumptions are not met in practice. We explored ``one-to-many'' scenarios in which one record in $\bm{X}_2$ has multiple plausible matches in $\bm{X}_1$, and showed how both \texttt{fabl} and \texttt{BRL} demonstrated undesirable qualities. Other issues arise under ``many-to-one'' scenarios, where one record in $\bm{X}_2$ has multiple
plausible matches in $\bm{X}_1$, and ``many-to-many'' scenarios in which there is duplication both across and within datasets. Tuning \texttt{fabl} for use in these scenarios is one potential avenue for
future work.

@@

\section{Appendix}
\label{sec:appendix}

@@

\bibliographystyle{jasa}
{
\footnotesize
\bibliography{biblio}
}

>>>>

In this paper, we propose the \emph{fast beta linkage} method (\texttt{fabl}), which extends a recent Bayesian Fellegi Sunter model for increased efficiency and scalability. Specifically, we relax the one-to-one matching requirement of the Beta Record Linkage method of \citep{sadinle_bayesian_2017} and propose independent priors over the matching space. This allows us to employ hashing techniques that hasten calculations and reduce the overall computational complexity.  In addition, we are able to complete pairwise record comparisons over large datasets through parallel computing and reduce memory costs through a new technique called \emph{storage efficient indexing}. Through simulation studies and a case study of homicides from the El Salvadoran Civil War, we show that our method has markedly increased speed with minimal loss of accuracy.
\end{abstract}

@@

	Record linkage is the task of identifying duplicate records across multiple data sources, often in the absence of a unique identifier \citep{christen_2012}. This is an increasingly important task in ``data cleaning,'' and is used for inferential and predictive analyses in fields such as statistics, computer science, machine learning, political science, economics, precision medicine, official statistics, and others. In this paper, we consider bipartite record linkage, which merges two databases together that contain duplications across but not within the respective databases.

	Many statistical record linkage methods are extensions of the seminal work of \cite{fellegi_theory_1969} and \cite{newcombe_automatic_1959}. Specifically, Fellegi and Sunter created comparison vectors for each pair of records in the data and independently classified those pairs as a match or a non-match using a likelihood ratio test. Recent work in the statistical literature has extended this approach for a wide variety of applications \citep{winkler1991application, fair2004generalized, wagner2014person, gill2003english}. Although the need to compute comparisons for all record pairs initially restricted this family of methods, recent work by \cite{enamorado2019using} used innovative hashing techniques to scale the approach to handle large datasets.
	% Possible citation for bipartite matching. 
	%\citep{fellegi_theory_1969, jaro1989, Winkler1988, belin_1995, larsen_2001, liseo_2011,  herzog2007data, gutman_bayesian_2013, sadinle_bayesian_2017}.

@@

	Alternatively, one can embed one-to-one matching requirements into the model specification itself \citep{gutman2013bayesian, liseo_2011}, at an additional computational cost. \cite{larsen2005} employed a Metropolis-Hastings algorithm to only allow sampling matches that respected one-to-one assumptions, but such algorithms exhibit slow mixing due to the combinatorial nature of the constrained matching space. \cite{fortunato_2010} used simulated annealing to target the space of matches permitted under the one-to-one constraint, but the method was so computationally intensive it could only be used to link databases of less than 100 records. Most recently, \cite{sadinle_bayesian_2017} proposed the \emph{Beta Record Linkage} model (\texttt{BRL}), using an innovative prior over the space of bipartite matchings to strictly enforce one-to-one reqiurements throughout his Gibbs sampler. Additionally, he introduced a class of loss functions that allows for a flexible estimation of the linkage structure, such that the modeller can weight the relative importance of false positives and false negatives, and identify records pairings to be be decided through clerical review. Although it was shown to work on larger tasks than previous one-to-one methods, \texttt{BRL} becomes slow when working with larger datasets, and unusable when the number of pairwise record comparisons becomes too large.

	In this paper, we propose \emph{fast beta linkage} (\texttt{fabl}), which extends the the Beta Record Linkage model for increased efficiency and scalability. Specifically, we relax the one-to-one matching requirement of \texttt{BRL} and propose independent priors over the matching space, creating a ``many-to-one'' model for record linkage. This allows us to (1) employ hashing techniques that hasten calculations and reduce computational costs; (2) compute the pairwise record comparisons over large datasets via parallel computing; and (3) reduce memory costs through our proposed method of \emph{storage efficient indexing}. Throughout, we argue that even in cases where a bipartite matching is desired, our proposed approach (with a simple post-processing procedure) provides equivalently accurate estimation of the linkage structure and other parameters, more information through which to asses model misspecification, and greatly enhanced speed. Open source software to use \texttt{fabl} in \texttt{R} is available through \href{https://github.com/briankundinger/parlrdev}{Github}.

	In what follows, Section~\ref{sec:review-of_prior-work} reviews the prior work of \cite{fellegi_theory_1969} and \cite{sadinle_bayesian_2017}. Section~\ref{sec:fast-beta-linkage}, proposes the \texttt{fabl} model, derives the Gibbs sampler for posterior inference, and provides the loss function used to calculate the Bayes estimate for the bipartite matching. 
	Section~\ref{sec:efficiency} introduces the hashing technique and storage efficient indexing used to increase the speed of calculations and the scale of the linkage tasks we can undertake. In addition, we prove the computational complexity of \texttt{fabl}. Sections~\ref{sec:simulations} and \ref{sec:case-studies} demonstrate the speed and accuracy of \texttt{fabl} through simulation studies and a case study of homicides from the El Salvadoran Civil War. Section~\ref{discussion} concludes our paper with a discussion of directions for future work.

@@

In practice $P(\cdot \mid \Delta_{ij} = 1)$ and $P(\cdot \mid \Delta_{ij} = 0)$ are unknown to the modeler, and thus must be estimated. Under the Fellegi Sunter model, we assume both that comparison vectors are independent given their matching status, and that the match status of the record pairs are independent. This allows us to model 
the comparison data using mixture models in the following way:
\begin{align}
	&\Gamma_{ij} = \gamma_{ij} \mid \Delta_{ij} = 1 \stackrel{iid}{\sim} \mathcal{M}(\bm{m}), \\
	&\Gamma_{ij} = \gamma_{ij} \mid \Delta_{ij} = 0  \stackrel{iid}{\sim} \mathcal{U}(\bm{u}), \notag\\
	& \Delta_{ij}   \stackrel{iid}{\sim} \text{Bernoulli}(p), \notag
\end{align}
where $p$ is the proportion of records that match, and $\bm{m}$ and $\bm{u}$ are vectors of parameters for the matching and nonmatching record pairs. These unknown parameters are often estimated using the EM algorithm \citep{Winkler1988}.

@@

Importantly, this independence means that we have relaxed the one-to-one requirement from \texttt{BRL}; our sampler does ensure that each record in \(\bm{X}_2\) can be matched to at most one record in \(\bm{X}_1\), but allows for the possibility that multiple records in \(\bm{X}_2\) match to the same record in \(\bm{X}_1\). This behavior may be desirable in its own right; \cite{Newcombe1963} was interested in matching birth records to marriage certificates, a situation in which such ``many-to-one'' matchings are reasonable. In situations when a bipartite estimate is desired, we can use a simple post processing procedure shown in  section~\ref{bayes-estimate} to resolve illegal matchings.

@@

With the likelihood in this form, we can derive the full conditional
\begin{align*}
	p(Z_j|\Gamma_{.j}, \Phi, \pi) &\propto p(\Gamma_{.j}| Z_j, \Phi) P(Z_j|\pi) \\
	&\propto \left(\sum_{i=1}^{n_1}w_{ij}\bm{1}_{z_j = i} + \bm{1}_{z_j = n_1 + 1}\right)\left(\pi\sum_{i=1}^{n_1}\frac{1}{n_1}\bm{1}_{z_j = i} + (1-\pi)\bm{1}_{z_j = n_1 + 1}\right) \\	
	&= \frac{\pi}{n_1}\sum_{i=1}^{n_1}w_{ij}\bm{1}_{z_j = i} + (1-\pi)\bm{1}_{z_j = n_1 + 1} \\
	&\implies Z_j^{(s+1)} | \Phi, \Gamma, \pi \propto \begin{cases} 
	\frac{\pi}{n_1}w_{ij}   & z_j \leq n_1; \\
	1-\pi &  z_j  = n_1 + 1 \\
\end{cases}
\end{align*}

@@

To increase the speed of our Gibbs sampler and expand the scale of the linkage tasks we can undertake, we propose hashing methods similar to those used by \cite{enamorado2019using} in the creation of \texttt{fastlink}, a fast and scalable implementation of the Fellegi Sunter method. The key insight is to recognize that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. To make this more precise, let $h_1, \ldots, h_P$ denote the unique agreement patterns, and collect these unique patterns in the set $\mathcal{P} = \{h_1, \ldots, h_P\}$. Here, 
$P = |\mathcal{P}|$ is the total number of unique agreement
patterns; observe that \(P\) is bounded above by \(\prod_{f=1}^F L_f\), and that this bound does not depend on \(n_1\) or \(n_2\). In this context, the integers $\{1, \ldots, P\}$ serve as \emph{hashed values} that encode the same information as the orignial vectors themselves. Whenever possible, we conduct calculations over these \(P\) agreement
patterns, instead of the typical \(n_1 \times n_2\) record pairs.

@@

In this paper, we have proposed \texttt{fabl}, a Bayesian record linkage method that scales to large data sets. Specifically, our proposed hashing method
and model assumptions allow for a linkage procedure whose computational complexity does not scale with the size of the larger dataset, making our approach particularly powerful in linking records when one datafile is substantially smaller than the other.

Throughout our paper, we have discussed independent pairwise matching assumptions, one-to-one matching assumptions, and when such assumptions are not met in practice. We explored ``one-to-many'' scenarios in which one record in $\bm{X}_2$ has multiple plausible matches in $\bm{X}_1$, and showed how both \texttt{fabl} and \texttt{BRL} demonstrated undesirable qualities. Other issues arise under ``many-to-one'' scenarios, where one record in $\bm{X}_2$ has multiple plausible matches in $\bm{X}_1$, and ``many-to-many'' scenarios in which there is duplication both across and within datasets. Extending \texttt{fabl} in such scenarios with a more robust linkage structure parameter $\bm{Z}$ is currently under investigation by the authors for future work.

To our knowledge, \texttt{fabl} is the first method within the Fellegi Sunter comparison vector approaches with an interpretable parameter for the rate of matching across datasets. This motivates extentions to \texttt{fabl} with additional parameters that allow the quality of information (the $\bm{m}$ and $\bm{u}$ parameters) and the rate of matching (the $\lambda$ parameter) to differ by subgroups in the linkage task. We see these types of heirarchical record linkage models as potential contributions to the growing algorithmic fairness literature, and useful to practitioners working with or studying marginalized populations.

%Tuning \texttt{fabl} for use in these scenarios is one potential avenue for
%future work.

@@

\bibliographystyle{jasa}
\bibliography{biblio}

\section*{Appendix}
\label{sec:appendix}

@@
